---
title: "Sheet 1: The Very Basics of R for Data Analysis"
author: "José Boue"
date: "2025-05-30"
output: 
  github_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1.1: What is R?

First and foremost, R is a piece of statistical software. R is useful because, among other things, it's free and open-source, which makes it easy to install and use anywhere. It is also highly modular, which in combination with the above makes it possible for R users to develop their own specialised packages for it. If you are having trouble doing something using built-in functions only, there is probably a package available somewhere that will let you do it with minimal difficulty.

What sets R apart from other software applications with similar purposes is that it is also a fully-fledged programming language. This can make using it seem daunting at first!

## 1.2: What isn't R?

R is not Excel. You can't expect to set up your tables just by clicking and dragging. All data manipulation in R must be performed by typing commands into the R console, or running them from script files.

R is not SPSS. You can't perform your analyses by navigating menus - you have to type all the code that defines your models and graphs out manually, although the RStudio interface will help you with autocomplete suggestions.

R is not Stata. Every line in Stata starts with a command, and every command has its own line. R is much less restrictive about how you can lay out your code, although that also means it takes longer to learn how to do it.

R is not a traditional programming language. R was designed from the ground up to be used for statistical analysis, which means that things which you take for granted about languages like C++ or Python (e.g. array indices starting from 0) have no guarantee of being true here.

R is not MATLAB. Frankly, MATLAB is awful. Its syntax is horrible, it hogs memory, and if it doesn't work it's nearly impossible to figure out what you did wrong. If, like me, you have used MATLAB before and hated it, I can reassure you that R is superior on all of those fronts.

## 1.3: Where does R keep your data?

R stores everything that you are currently using in the **environment**. Unlike in Stata or SPSS, you can have more than one data set active at the same time!

However, the environment is not just for data. It can store any R object, including **variables** and **functions**.

## 1.4: How do you tell R to store something?

R uses very similar syntax for creating all kinds of objects. The key is the **assignment** operator, written "<-". In most programming languages, you declare the value of a variable using a single equals sign "=". That is not the case in R, because R uses assignment to define functions as well as variables, and using the equals sign to define a function would look a little odd.

The assignment operator points towards the left to help you visualise the program flow. When you define a variable, its name goes on the left and its value goes on the right. You are **assigning** the value to the name, not the other way around. The code could look like this:

```{r examplevar}
x <- 1
```

If you run this code, you will see a variable called x appear in the environment (in the upper right corner of the screen), and you can see that its value is 1. So far, so good. If you hover over it, you will see that its type is **numeric**. We could have assigned x a string as its value, in which case its type would be **character**, or a Boolean (TRUE or FALSE) in which case its type would be **logical**. We don't want to overwrite x as we might need to use it later, so we can assign these values to two new variables instead.

```{r vartypes}
s <- "Hello World!"
b <- TRUE
```

What do you do if you decide you don't need a variable any more and want to get rid of it entirely, rather than overwrite its value with something else? There is a special function for that: **rm()**, which is short for **remove**. When it is given the name of any R object currently in the environment, it does just that. We will see more advanced uses of rm() in the appendix.

```{r removal}
mistake <- "This variable was a mistake."
rm(mistake)
exists("mistake")
```

Note the use of the exists() function to determine if an object of the provided name is present in the environment.

## 1.5: How do you import data from a file?

Most of the time, you will be given data for analysis in the form of a text (**.txt**) or comma-separated (**.csv**) file. Excel files (.xlsx) are not good for this because Excel's auto-formatting can introduce errors, so if someone gives you data in an Excel file do not try to import it. Instead, dispose of the file and kindly inform whoever gave it to you not to do that. Excel files also have a hard limit of 1,048,576 rows, which only seems like a lot to anyone who has never worked with big data. Big data is BIG.

There are two ways to import a data file into R. The first is by clicking the **"import dataset"** button just above the environment window, choosing the file type you want, and then navigating to the file in the Explorer window. This is the most "SPSS-like" thing that R lets you do.

The second way is by using the console. R has built-in functions for importing files, which are called read.table and read.csv. These functions have multiple arguments, but the most important ones that you need to know about for now are the file path, and whether or not the first row of the file should be read as a header.

When calling read.table or read.csv, the file path must be written as a string (enclosed by quotes). In order to save your file in the environment, you must assign it a name. Here is an example of what the code could look like:

```{r examplereadtxt, eval=FALSE}
my_table <- read.table("C:/Users/me/Documents/example.txt", header=TRUE)
```

If any arguments of an R function are not specified, they take their default values (provided such values exist, if they don't it gives you an error). We needed to specify that the first row should be read as a header (for data you work with, this will almost always be the case) because by default it is FALSE for read.table. For some reason, for read.csv it is TRUE by default, so you don't need to include it. Be careful not to get caught out by this!

This seems like much more of a fuss than simply using the menu, but you will need to use these commands when importing multiple data sets in sequence (or else you'll get very bored of opening the menu again and again). We'll learn more about how you can eliminate this kind of busy work when we look at control flow.

A word of caution: saving data in the environment is not the same as saving it as a file on your computer. If you want to do the latter so you can easily load it again where you left off, you need to click the floppy disk icon above the environment and save it as an .RData file. Unlike .csv or .txt files, .RData files can contain multiple data sets, as well as any variables or functions you've created in order to manipulate them. You can load an .RData file into your environment by clicking the folder icon next to the save icon.

There are separate icons for saving and loading above the source window, which is where your R scripts live. Just be sure to remember that saving or loading an R script file (with extension .R) will not do the same for its associated data!

## 1.6: How do you manipulate your data?

Once you have imported your file, it is stored in R as an object known as a **data frame**. A data frame is basically a fancier version of a **matrix**, which itself is a two-dimensional version of a **vector**. What is a vector, you might ask?

A single-valued variable, such as x from before, is known as a **scalar**. Vectors in R are formed using a special function, **c()**, which **concatenates** (hence the name) any number of scalars of the same type (numeric, character, or logical) into a vector. Here's how you might do it:

```{r vec1}
y <- 2
z <- 3
my_vector <- c(x,y,z)
my_vector
```

Note that since x is saved in the environment, you don't have to define it again. We used x here to show this functionality, but we could have just as easily used the numeric constant 1 instead; c() accepts numbers, strings or logical values (TRUE or FALSE) even if they haven't been bound to a variable name. Let's see some examples using the other variables we defined earlier:

```{r vec2}
c(b, FALSE)
c(s, "This is a string.")
```

In a data frame, each column is a named vector. You can manipulate any column of a data frame using the same methods for manipulating vectors. This leads into our next question...

## 1.7: How do you manipulate vectors?

The first thing you need to know when working with vectors is that c() is not just for combining scalars. It can also combine vectors with scalars, or even with other vectors. Try it:

```{r vec3}
c(my_vector,x)
c(my_vector,my_vector)
```

You can see that we used c() to duplicate our vector by typing its name twice. This seems a bit inefficient, so you're probably wondering if there's a better way to do it. Luckily, there is: the **rep()** function. In its most basic form, this function has two arguments: the first is a scalar or vector, and the second is a number that tells R how many times to repeat it. See that it gives you the same result as before:

```{r vec4}
rep(my_vector,2)
```

The second argument need not be a constant, either. This is also allowed:

```{r vec5}
rep(my_vector,y)
```

We can do some slightly more advanced things with rep() that we'll look at in the appendix.

Aside from rep(), there is another common method for quickly generating vectors of arbitrary length. If you want to make a numeric sequence rather than a simple repetition of the same few elements, you can use the **:** operator. Simply typing n:m, where n and m are any integers, makes a vector of all integers between n and m inclusive (if n is more than m it will output them in reverse).

There is a function that does the same thing as : but allows for more options called **seq()**, which we will also look at in the appendix.

## 1.8: Using mathematical operators with vectors

Like most programming languages, R can be used as a fancy calculator. If you type a mathematical expression into the console, it will give you the answer. The standard mathematical operators (+, -, \*, /) are not just usable with numbers, though. They can also be applied to vectors, which is something that traditional languages will not let you do (without a for loop, that is). Operators are applied **element-wise**, which means that performing an operation on two vectors of the same length is equivalent to applying it on each pair of values in the same position in sequence.

```{r vecmath1}
rep(1,5)+rep(2,5)
```

If the vectors are not the same length, it can still work; however the length of one vector must be an exact multiple of the length of the other. What happens in this case is that the smaller vector is repeated to match the length of the other before the operation is carried out. This means that you can add a number to every element of a vector at once simply by typing the following:

```{r vecmath2}
my_vector+1
```

Needless to say, most programming languages will throw a hissy fit if you try to do anything like this with two objects that have different lengths. Subtraction, multiplication and division work the same way.

## 1.9: Working with data frames

A data frame is made of multiple column vectors of the same length bound together, each of which represents a variable. Unlike in a matrix, these columns can have different types, which is useful if you're working with a data set that has both numeric and character variables (for example, a list of patient IDs with associated ICD-10 codes). Remember, however, that you can't put values of different types in the same column.

R has a special syntax for referencing a particular column of a data frame. In order to do this, write the name of the data frame followed by a dollar sign **$** and the name of the column. Any method that works on vectors can be applied to such an object. You can add a new empty column with your preferred name to an existing data frame simply by doing this:

```{r frame1, eval=FALSE}
my_table$new_column <- rep(0)
```

Note that you don't need to specify how long the column vector should be, as R makes it the same length as the other columns automatically. When I make empty columns like this I prefer to fill them with zeroes rather than having them be truly empty (R represents this with the value **NA**) because having a column without a defined type can cause problems when trying to fill it in with actual data.

An alternate method for adding an existing vector to a data frame as a new variable is using the **cbind()** function. cbind() takes one or more column vectors and/or data frames as arguments and binds them together sequentially. However, using cbind() does not allow you to name your vectors as you add them: you have to do it manually later. We'll explain how to bind and name at the same time once we get to square brackets.

But what if you don't want to add a new column, but a new row instead? There is another function, naturally named **rbind()**, that does just that. Your new row can't be saved as a vector unless all its entries have the same type, though: you must create it either as a **list** or as a data frame with 1 row.

A list is to a vector as a data frame is to a matrix. While vectors are limited to storing scalars of a single type, lists can store **anything** as entries, including vectors, matrices, data frames or even other lists. Many statistical models you will run return their outputs as lists. You don't need to know much about lists right now, but keep them in mind for later.

## 1.10: Assembling data frames from scratch

Let's say you have a bunch of data vectors that you want to assemble into a data frame. Using cbind() on them all would produce a matrix, not a data frame, as cbind() requires at least one of its arguments to be a data frame for the final product to be one. If you mix numeric and character vectors, R still won't "guess" that the result you want is a data frame; instead it will coerce the numeric vectors into characters.

Thankfully, there is a function that allows you to create your own data frames, which is conveniently called **data.frame()**. Calling this function with no arguments produces a data frame with zero rows and zero columns, which seems odd but you can still add data to it later to give it a defined size. This will be helpful when working with loops. Each argument of data.frame() is a column vector, which you can assign a name to. The syntax for doing so looks like this:

```{r frame2, eval=FALSE}
my_df <- data.frame("variable_1"=vector1, "variable_2"=vector2, "variable_3"=vector3)
```

You can also use this method with the $ operator to extract specific columns from an existing data frame to make a new one, although the **subset()** function (covered in the appendix) and the square brackets can do the same thing more elegantly.

## 1.11: Using square brackets

You can also reference columns, rows, or even single entries of a data frame using the **square brackets []**. 

In one dimension, the square brackets are used to reference a single entry of a vector. Unlike in standard programming languages, the indices of vectors start at 1 rather than 0. You also can't use negative indices to reference entries starting from the end (for example, in Python L[-1] would return the last entry of a list L). In R, putting the minus sign **-** in front of an index does something rather different: it actually drops that entry from the vector, although it doesn't edit the original. It makes a copy of it without that entry, which it won't save unless you assign it a name. If you want to edit a vector to remove the first entry, you can overwrite it with the new copy like this:

```{r sqbrackets1, eval=FALSE}
my_vector <- my_vector[-1]
```

You can reference multiple entries of a vector by putting a vector of their indices inside the square brackets. You can define these using c(), the : operator, or any other method you like. Both lines below produce the same result:

```{r sqbrackets2, eval=FALSE}
my_vector[c(1,2,3)]
my_vector[1:3]
```

Putting a - in front of the vector inside the square brackets will remove **all** the entries it references. Instead of numeric indices, you can also use **logical** vectors as indices. Putting a logical vector (which must be of the same length as the vector you are referencing) inside the square brackets will return only those entries for which the value in the same position in the logical vector is TRUE. This may not look it at first, but it's an extremely powerful method. The reason for this is because any logical statement involving vectors will generate a vector of logical values, which can be used as an indexing vector. We'll cover how to use logical statements soon.

Using square brackets to reference values in a matrix or data frame is similar, only you now have to specify two arguments inside the brackets separated by a comma. If you want only specific columns or rows, you leave one argument blank, but the comma still needs to be included. This gives you a way to reference variables without typing their full names out, although you need to remember the column number and be careful with moving data around. For our example data frame, these two lines output the same values:

```{r sqbrackets3, eval=FALSE}
my_df$variable1
my_df[,1]
```

Of course, you can also use the minus sign **-** to quickly drop rows or columns that you don't need.

Another thing you can use square brackets to do easily is renaming variables. There is a function called **names()** that, when applied to a matrix or data frame, returns a vector containing all the column names in order. To rename, say, the 10th column, all you have to do is type the following:

```{r sqbrackets4, eval=FALSE}
names(my_df)[10] <- "better_name"
```

There is also a **rownames()** function, which is less useful but can still help sometimes.

## 1.12: Logical operators

The basic logical operators in R **(==, !=, <, >, <=, >=, |, &, !)** are mostly the same as in other programming languages. Testing for membership in a set is done with the operator **%in%**, which has percentage signs around it to differentiate it from the "in" used in **for** loops. Like standard mathematical operators, logical operators are applied element-wise to vectors, which means using < and > with vectors actually makes sense. For example, the following line will return a logical vector describing whether each entry of my_vector is greater than 1:

```{r logic1}
my_vector>1
```

This means that the following code will keep only the values of my_vector that are greater than 1:

```{r logic2}
my_vector[my_vector>1]
```

What if you're not interested in the magnitude of the values that are greater than 1, but simply want a vector of their positions? This is done using the **which()** function. This function takes a logical vector as its argument, and returns the positions of which elements are TRUE. Let's not use my_vector to illustrate this, because its positions and values are the same.

```{r logic3}
my_fancy_vector <- c(4,2,0,6,9)
which(my_fancy_vector>3)
```

Other useful logical functions include **is.na()**, which tells you the positions of any values that are missing, and **is.finite()**, which tells you which values are numbers (NA, NaN, Inf, and -Inf all return FALSE, if you don't know what these values mean consult the appendix). A lot of the time, NA values will mess up your calculations, so when formulating your logical statements you will have to add an is.na() call, like this:

```{r logic 4,eval=FALSE}
with(my_df, variable_2[variable_1==x&!is.na(variable_1)])
```

Note the use of **&** (logical AND) and **!** (logical NOT) here. You can use these as well as **|** (logical OR) to make compound logical statements, saving you from having to pass logical functions multiple times. The **with()** function is also very useful, as calling it with a data frame as the first argument and an expression to be evaluated as the second allows you to reference its columns by name in the expression without needing the dollar sign operator. If you find yourself typing the name of a data frame out very often, consider using with() to save you some space.

## 1.13: Basic control flow

Like in standard programming languages, R's control flow primarily uses **if** statements and **loops**. As we have seen, loops are not needed for simple vector operations in R even though they would be in other languages, but that does not mean R loops are not useful.

The basic syntax is the same for all control flow commands. It consists of the command, followed by a statement in brackets that determines the conditions under which the desired effect will happen, followed by the main definition of what that effect will actually do in **curly brackets {}**, which is called the **"body"**. Let's see what the simplest version of this, the if statement, looks like:

```{r if, eval=FALSE}
if(is.na(x)){x <- 0}
```

What this code does should be easy to understand: it checks if the value of a variable x is missing, and if it is, it replaces it with 0. In an if statement, we can put any logical statement in the first pair of brackets. 

Naturally, an if statement can be accompanied by an **else** clause, which requires no condition. Like in some other languages, the way to make multiple-choice if statements is by chaining the else clause into another if statement. There is no direct equivalent of Python's "elif".

```{r elif, eval=FALSE}
if(x==1){print("The value of x is 1.")}
else if(x==0){print("The value of x is 0.")}
else{print("x is not a binary variable.")}
```

R also has a function called **ifelse()** that is designed to replace very simple if-else statements for which writing them out in full would be cumbersome. This function has three arguments: the first is a logical statement, the second is what the function will output if that statement is TRUE, and the third is the output if that statement is FALSE. In effect, the two lines below are the same:

```{r fn_ifelse, eval=FALSE}
ifelse(condition, "Yes", "No")

if(condition){print("Yes")} else{print("No")}
```

Loops are similar to if statements, only instead of evaluating their condition once they do it multiple times in succession. There are three types of loops in R, **for**, **while**, and **repeat**.

The **for** loop is the most intuitive to use. Its condition is not a logical statement, but is of the form **i in v** where i is a dummy variable and v is a vector of your choice. For instance, if we want to expand our if-else statement above to deal with a variable that can take any integer value from 0 to 9 rather than just 0 or 1, we can use a for loop:

```{r forloop, eval=FALSE}
for(i in 0:9){
  if(x==i){print(str_glue("The value of x is {i}."))}
}
```

The **str_glue()** function is a very useful function for string manipulation which is part of the **stringr** package. We'll look more closely at it and its associated package in the appendix.

Typically, when manipulating data you will only need to use for loops. While and repeat loops are designed to be used when you don't know in advance how many iterations you will require, which is rare when you're dealing with data as you'll always know how big your tables are. For that reason, these have been relegated to the appendix.

## 1.14: Custom functions

R and its packages have many functions that can do almost anything, but at some point you will need to write your own. Writing functions can save time, as you can call a function anywhere in your code without having to write a complicated expression in full again. Function definitions use a similar syntax to control flow statements, but they must be assigned a name like variables. The basic structure looks like this:

```{r function1, eval=FALSE}
my_function <- function(args){body}
```

The first pair of brackets contains all the function's arguments, which like in a for loop are just dummy variables that can be used again in the function body. The function's body determines what its output will be. In order to make sure a function outputs only a single value, use the **return()** function. Once the return() function is called, the main function will terminate, and all body code below the return() will be ignored. Other functions like **print()** do not work this way, so you can easily make a function that prints a succession of values using a loop. Note that unlike in some other languages, print() does not automatically turn its argument into a string.

```{r function2, eval=FALSE}
loop_function <- function(n){
  for(i in 1:n){print(i)}
}
```

The above function takes an integer n as its argument, and prints all integers from 1 to n on separate lines. This function has no error handling, so giving it a bad input can cause problems. When writing your own functions that will be used by other people, be sure to think about whether any possible inputs might result in unintended behaviour and include code to handle them.

## 1.15: Where should you go if you have further questions?

R has an inbuilt method for helping with unfamiliar code. If you ever see a function you don't recognise or understand, you can type ? followed by the function name in the console and R will automatically open the help page for that function. Default functions will have very thorough documentation, but due to R's open-source nature if you are using specialised packages you are at the mercy of how dense or sparse the author's explanation is. In cases where the official documentation is not enough, you can do the classic moves of either resorting to StackOverflow or asking an AI to explain it to you, but I believe the best course of action is always to play around with the code until you figure it out. If you are careful not to overwrite your existing data, messing up won't cost you much other than your time. Good luck!