---
title: "Sheet 2: Statistical tests, models, and charts in R"
author: "José Boue"
date: "2025-06-23"
output: 
  github_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

All functions in this document are assumed to require the `stats` package, which is built-in to current versions of R and doesn't need to be manually installed.

# 2.1: Tests

If a function has an argument `alternative`, this can be `two.sided`, `less`, or `greater`.

## 2.1.1: *z*-test

```{r ztest}
pnorm(mu, mean=mu_0, sd=s) #one-sided
```

## 2.1.2: *t*-test

```{r ttest}
t.test(x, alternative, mu) #one-sample
t.test(x, y, alternative, mu_diff) #two-sample
t.test(x, y, alternative, mu_diff, paired=TRUE) #two-sample, paired
```

## 2.1.3: Chi-squared test

```{r chisq}
chisq.test(x, p) #one-way table with probabilities
chisq.test(x, y) #two-way table as two vectors
chisq.test(M) #two-way table as a matrix
```

## 2.1.4: *F*-test

```{r ftest}
pf(v, df1, df2) #one-sided
```

## 2.1.5: Fisher's exact test

```{r fisher}
fisher.test(x, y, alternative) #inputs are two vectors
fisher.test(M, alternative) #input is a matrix
```
## 2.1.6: Kruskal-Wallis test

```{r kruskal}
kruskal.test(x, g) #vector with groupings
kruskal.test(L) #list of vectors
```

## 2.1.7: Pearson correlation

```{r pearson}
cor.test(x, y, alternative, method="pearson")
```

## 2.1.8: Spearman's rank correlation

```{r spearman}
cor.test(x, y, alternative, method="spearman")
```

## 2.1.9: Wilcoxon signed-rank test

```{r wilcox}
wilcox.test(x, mu, alternative) #one-sample test
wilcox.test(x, y, mu, alternative) #two-sample test
wilcox.test(x, y, mu, alternative, paired=TRUE) #two-sample paired test
```

## 2.1.10: Kolmogorov-Smirnov test

```{r kstest}
ks.test(x, y, alternative) #two-sample test
ks.test(x, cdf, alternative) #one-sample test
```

# 2.2: Models

## 2.2.1: Linear regression

```{r linreg}
lm(y ~ ..., data)
```

## 2.2.2: Poisson regression

```{r poisson}
glm(y ~ ..., family="poisson", data)
```

### 2.2.2a: Conditional Poisson regression (untested)

Requires `acp` package.

```{r cpoisson}
acp(y ~ ..., data, p, q, family="poisson")
```

## 2.2.3: Logistic regression

```{r logit}
glm(y ~ ..., family="binomial", data)
```

### 2.2.3a: Conditional logistic regression

Requires `survival` package.

```{r clogit}
clogit(y ~ ..., data)
```

## 2.2.4: Cox regression

Requires `survival` package.

```{r coxreg}
coxph(Surv(t,y) ~ ..., data)
```

## 2.2.5: Self-controlled case series

Requires `SCCS` package.

```{r sccs}
#Basic model, dose-independent exposure
standardsccs(event ~ adrug[,1], indiv, astart, aend, aevent, adrug, aedrug, data)
#Dose-dependent exposure
standardsccs(event ~ adrug_i, indiv, start, aend, aevent, adrug, aedrug, data)
#Age effect
standardsccs(event ~ adrug[,1] + age, indiv, astart, aend, adrug, aedrug, agegrp, dob, data)
#Seasonal effect
standardsccs(event ~ adrug[,1] + season, indiv, astart, aend, adrug, aedrug, seasongrp, dob, data)
#With washouts and/or pre-exposures
standardsccs(event ~ adrug[,1], indiv, astart, aend, aevent, adrug, aedrug, expogrp, washout, data)
```

## 2.2.6: Accelerated failure time

Requires `survival` or `eha` package.

```{r aft}
survreg(Surv(t,y) ~ ..., data, id, dist) #survival package version
aftreg(Surv(t,y) ~ ..., data, id, dist) #eha package version
```

## 2.2.7: LOESS regression

```{r loessreg}
loess(y ~ ..., data)
```

# 2.3: Charts

No purely cosmetic arguments are included here, only ones that affect how the data is interpreted.

## 2.3.1: Scatterplot

```{r scatter}
plot(x, y, type)
```

## 2.3.2: Bar plot

```{r barchart}
barplot(height=v, width) #simple bar plot, v is a vector
barplot(height=M, width) #juxtaposed bar plot, M is a matrix
barplot(height=M, width, beside=FALSE) #stacked bar plot
```

## 2.3.3: Box plot

```{r box}
boxplot(y ~ ..., data)
```

## 2.3.4: Histogram

```{r histogram}
hist(x, breaks, right)
```

## 2.3.5: Pie chart

```{r piechart}
pie(x)
```

## 2.3.6: Radar chart

```{r radar}
stars(x, full, locations)
```

## 2.3.7: Q-Q plot

```{r qq}
qqnorm(y) #Normal distribution Q-Q plot
qqplot(x, y) #Generic Q-Q plot
```

## 2.3.8: Forest plot

Requires `forestploter` package.

```{r forest}
plot(forest(data, est, lower, upper, sizes, ci_column, ref_line))
```

## 2.3.9: Kaplan-Meier plot

Requires `survival` package.

```{r kmplot}
plot(survfit(y ~ x, data))
```

## 2.3.10: Choropleth map

Requires `sf` package. Other packages optional.

```{r choropleth}
plot(shapefile[,"var_name"])
```

## 2.3.11: Cartogram

Requires `cartogram` package. Other packages optional.

```{r cartograms}
plot(cartogram_cont(shapefile, weight)) #contiguous cartogram (shape-distorting)
plot(cartogram_ncont(shapefile, weight)) #non-contiguous cartogram (shape-preserving, but leaves gaps)
plot(cartogram_dorling(shapefile, weight)) #Dorling cartogram (represents regions as circles)
```