---
title: "Session 1a: Basics of R (Appendices)"
author: "José Boue"
date: "2025-10-23"
output: 
  github_document:
    toc: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Appendix 1: Extra arguments for `read.table()` and `read.csv()`

Sometimes, when working with categorical data you will want the relevant variables to imported into R as **factors** rather than basic strings. You can do this by setting the argument `stringsAsFactors` to `TRUE` (by default it is `FALSE`). A factor is simply R's name for a categorical variable, where there are only a set number of values (levels) it can take that may or may not be ordered. When you import a variable as a factor, all values that R sees are turned into its levels. Be careful, as if a particular value is not present in the table R won't add it as a level.

The `sep` argument is what determines the character that R reads as separating data values in the file. By default, it is empty in `read.table()`, which means any whitespace character (tab, space, newline, carriage return) will be read as a separator. You can set it to a comma in order to read .csv files without using `read.csv()`, or a tab ("\\t") to read tab-delimited files. The `read.csv()` function naturally has the comma as its default separator. In some European countries the comma is used in place of the decimal point and the semicolon as a separator, so be sure to change the `sep` argument accordingly if you're working with files from Europe.

## Appendix 2: Writing to a .txt or .csv file

Like `read.table()` and `read.csv()`, there also exist `write.table()` and `write.csv()` functions. These functions take a data frame and write it to a specified file path. Other arguments include whether character columns are quoted, the separator to use (`write.table()` only), whether row and column names are to be included in the file, and the string to use to denote missing values.

## Appendix 3: Reading files directly from the internet

In addition to ordinary file paths, `read.table()` and `read.csv()` also accept URLs. This allows you to import a table hosted online (as a text file, ordinary web pages won't work) without having to download it to your computer. Using this trick can make it easier for other people to run your code, although it has quite a severe restriction in that any file you use it on has to be publicly accessible.

## Appendix 4: Extra arguments for `rep()`

The `rep()` function actually has three possible arguments that can come after the vector you want to replicate. The first is `times`, which just determines how many times you repeat it. By default, this is the argument that any number coming after the vector will be fed into. If you want to access the other two arguments, you need to spell them out. The other two arguments are `each`, which repeats each individual element of the vector for the specified number of times in sequence, and `length.out`, which repeats the whole vector until a specified length is reached (it can cut it off in order to do so). Let's see what the code looks like:

```{r reps}
my_vector <- c(1, 2, 3)
rep(my_vector, each=2)
rep(my_vector, length.out=7)
```

## Appendix 5: The `seq()` function

Like the `:` operator, `seq()` defines a vector that contains a sequence of numbers. It has four arguments: `from`, `to`, `by`, and `length.out`. The first two determine where the sequence starts and ends, while `by` determines the difference between successive members of the sequence and `length.out` the length the sequence should have. Only one of the last two arguments should be specified, as the value of one uniquely determines the other. There are variants of `seq()` that require fewer arguments, such as `seq_along()`, which when given a vector returns a sequence from 1 to its length. This is not exactly the same as using `1:length(v)`, because `1:0` produces a vector equal to `c(1, 0)` instead of a zero-length vector, so `seq_along()` is useful in this case for avoiding unintended behaviour.

## Appendix 6: The `subset()` function

This function can be applied to vectors, matrices, or data frames. In its simplest form, it takes as its second argument a logical expression that indicates which elements to keep. This is mostly equivalent to using square brackets with a logical expression inside, although unlike that method `subset()` treats missing entries as `FALSE`. If you use `subset()` on a matrix or data frame, you can specify another argument, `select`, that lets you choose which columns to keep by referencing their names. You can put a `-` in front to drop them instead.

## Appendix 7: Lists

Lists may seem intimidating, but they're not actually that difficult to use. They can take up a lot of space, however. The way to generate a list is much the same as for a vector, only instead of `c()` you use the `list()` function. You can also name each element of your list like you would do with a data frame. Appending an object to a list is done using the `c()` function, but you need to be careful when doing this as mixing lists and vectors can lead to unintended behaviour. For best results, turn the object you want to add into a length-1 list before using `c()`.

Extracting single elements from a list is done using the **double square brackets** `[[]]`. You can use the single square brackets as well, but this creates a list of length 1 that contains the object you want rather than extracting the object itself. If you want to extract the first element of a vector which is itself the first element of a list, the syntax would look like this:

```{r listvec, eval=FALSE}
my_list[[1]][1]
```

If your list elements are named, you can also use the dollar sign operator `$` to reference them like you would with a data frame, and even chain together multiple dollar sign expressions if your lists are nested:

```{r nestedlist, eval=FALSE}
#element is inside list2 which in turn is inside list1
list1$list2$element
```

Conversely, you can actually use the double square brackets with data frames as well because a data frame is essentially a two-dimensional list, but this is usually not necessary.

## Appendix 8: `Inf` and `NaN`

If you have used another programming language before, you probably know what `Inf` and `NaN` are and the difference between them. If not, don't worry. `Inf` simply means **"infinity"**, which often is not actually infinite but shorthand for "bigger than any number R is capable of calculating". For example, attempting to divide a number by zero will return `Inf`, but you also might get `(0, Inf)` as a confidence interval if your statistical model takes too long to converge. `-Inf` is the same but negative. `NaN` stands for **"Not a Number"**, which means that if a calculation returns it the answer likely doesn't even exist. For example, attempting to divide 0 by 0 or taking the tangent of $\pi/2$ will produce `NaN`. R will give you a warning message whenever `NaN` shows up, as when this happens something has usually gone badly wrong.

While calculations involving `NaN` are inadvisable to perform, `Inf` and `-Inf` are relatively well-behaved and can be used in ordinary code without causing issues. For example, the `cut()` function will convert a numeric vector into a factor by binning, where the bins are specified using the `breaks` argument. If you want the first or last bin to be open-ended (or both!), which is common for certain variables such as age, you can specify Inf as the upper limit or -Inf as the lower one. You may not expect that to work, but it does!

## Appendix 9: The `stringr` package

Unlike traditional programming languages, base R does not have an easy way to manipulate strings. The most basic objects that R is designed to work with are vectors and matrices which are typically numeric, so it makes up for its strength in dealing with these with a weakness in dealing with strings. The `stringr` package is designed to address this issue by introducing simple functions for manipulating strings, the most useful of which in my opinion is `str_glue()`. 

This function can concatenate multiple input strings into a new string rather than a vector, and adding the `.sep` argument allows you to choose a separator to place in between them. But the real magic of `str_glue()` is its ability to insert variable values directly into strings. If you enclose the name of a variable with curly brackets `{}`, `str_glue()` will render it as that variable's value. Therefore, you can change an input or output string simply by changing a variable rather than having to fiddle around with the string itself.

Base R actually has a function called `sprintf()` which can do the same thing, but it's carried over wholesale from C and hence has extremely fiddly and un-R-like syntax. I would not advise using this function unless you are trying to minimise the amount of package dependencies in your code, which can be relevant sometimes.

There are many other nice functions in the `stringr` package, but most of them make use of **regular expressions**. This is a topic which is too big and too complex to fit in this appendix, so expect to see an entire sheet dealing with it.

## Appendix 10: Advanced uses of `rm()`

Although `rm()` was one of the earliest things you encountered in the main document, its more advanced options have been left relatively late here, as you will need to read the preceding appendices to understand them. If you want to remove multiple objects at once, the most basic way to do this is to pass a vector of their names. There are two ways to do this:

```{r rm1, eval=FALSE}
rm(var1, var2, var3)
rm(list=c(var1, var2, var3))
```

Note that you if you're not writing out the names of the variables in full, you will need to use the `list` argument rather than the default. The `list` argument is useful for removing a pre-defined vector of variables to save time writing them out.

If you have a lot of variables you need to get rid of, or if their names are quite long, any method that involves listing them all in full is inconvenient. You can make life easier for yourself using another special function, `ls()`. When run with no argument, this returns a vector of the names of all objects in the environment, so the following code will clear it completely:

```{r rm2, eval=FALSE}
rm(list=ls())
```

You can do the same thing by clicking the broom icon above the environment, which is better because it asks you to confirm before deleting everything. Also, putting the above line in your actual code file is a very bad idea, as it can give anyone else who runs your code a very nasty surprise if they're not expecting all their data to get nuked without warning. If you must run it, do it from the console.

However, `ls()` is very powerful when used with a bit more finesse. It has an argument, `pattern`, which can take an ordinary string or, if you're feeling brave, a regular expression. To show you how it works, the following code will remove all objects that have the substring "var" in their names, achieving the same effect as the basic code you saw earlier:

```{r rm3, eval=FALSE}
rm(list=ls(pattern="var"))
```

Of course, you need to be smart with naming your objects if you want to make full use of this.

## Appendix 11: if-else versus ifelse()

I mentioned earlier that these two styles of coding a simple conditional statement have some subtle differences in their behaviour, which are not usually seen but can can cause very cryptic errors when they do show up. The ordinary if-else statement requires its condition to be of length 1, or else it will throw an error. The ifelse() function does not have this restriction: its output always has the same length as its condition, which can be more than 1. However, it applies its condition element-wise to its two possible outputs, meaning that neither of them can be longer than its condition, and if they are shorter they will be recycled. It's quite hard to explain exactly how this works, so here is some code illustrating it:

```{r ifelse}
ifelse(c(TRUE, FALSE, TRUE, FALSE),
       c("work it", "make it", "do it", "makes us"),
       c("harder", "better", "faster", "stronger"))
```

## Appendix 12: While and repeat loops

**While** loops are very close to if statements in their syntax. They also use a logical expression as their condition, but can evaluate it more than once. If the condition is `TRUE`, the loop will execute the code contained in its body, then after this has finished it will check the condition again. If it is still `TRUE`, the body will execute again, otherwise the loop will terminate. Unlike for loops, while loops can potentially run forever without halting, which means they require more caution to use. In order to terminate a while loop, you will need to either include code in the body that can set the condition to `FALSE`, or use the `break` command. This command terminates the loop it's in, and passes to the next line of code below it. Here is an example of a while loop that uses an if statement to decide whether to break:

```{r whileloop, eval=FALSE}
while(n>0){
  if(n%%1==0){
    n <- n-1
  } else break
}
```

This code is designed to take a positive integer n and decrement it by 1 repeatedly until it reaches 0. If n is not an integer, the loop will break instead of passing an invalid input.

A **repeat** loop is identical to a while loop, but with no condition: the only way to terminate it is by manually breaking it. If it is part of a function, you can break it with a `return()` command. You can achieve the same effect as a repeat loop by writing `while(TRUE)`.

## Appendix 13: Unfulfilled promises

So far, it has been stated that an empty data frame can easily be filled using a loop, the `read.table()` and `read.csv()` functions can be looped to import multiple files without much effort, and that `str_glue()` is excellent for modifying input strings. No examples for any of these have been given, because it is best to demonstrate them all together. Consider a folder called big_folder containing files named file_1.csv, file_2.csv, and so on up to file_100.csv. Importing all of these files and combining them into a single data frame, assuming that they all have the same number of columns, is not nearly as onerous as it would appear. All that is needed is to create an empty data frame, then write a loop that binds each file onto it in succession:

```{r grandfinale, eval=FALSE}
my_df <- data.frame()
for(i in 1:100){
  my_df <- rbind(my_df, read.csv(str_glue("C:/Users/me/Documents/big_folder/file_{i}.csv")))
}
```

## Appendix 14: Anonymous functions and the `apply()` family

If you know about lambda expressions in languages like Python, you will remember that their main use is to allow functions to be used without naming them, which can save a bit of space. In R, the standard method of defining functions can be used the same way. If you have a function that takes the name of another function as one of its arguments, you need not define that function outside it: you can simply write the definition inline, and it will work. The most common functions you will use this with are the `apply()` family, which let you (depending on which `apply()` function you use) take a vector, list or even a matrix and apply a function element-wise to it that doesn't normally work that way without going to the trouble of writing a loop. R has quite a few other useful higher-order functions (functions that take other functions as arguments), which may be covered on a future sheet.

## Appendix 15: Dates

There are no references to dates in the main document, but I figure they're important enough to include here anyway. Under the hood, R stores a date as a number of days since the 1st of January 1970, with earlier dates than that being negative. You can choose which format R displays dates in, but the preferred one is yyyy-mm-dd. You can perform some basic mathematical operations with dates. Adding a number to a date produces the date that number of days later, and subtracting it produces the date that number of days earlier. You can also subtract one date from another to produce an object known as a **difftime**. A difftime is simply a number of days with no index date, meaning that you can now add them together and even multiply and divide them by numeric vectors. Most functions won't accept difftimes or dates when they expect numbers, though, so if you're using time-series data you may need to convert them to standard numeric objects before running certain models.