---
title: "R Club Introduction"
author: "José Boue"
date: "2025-10-23"
output: 
  powerpoint_presentation:
    reference_doc: Shapes.pptx
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introducing the Team

- José Boue: <jose.boue@phc.ox.ac.uk>
- Nava Gottlieb: <nava.gottlieb@phc.ox.ac.uk>
- Jess Schiff: <jessica.schiff@phc.ox.ac.uk>
- Milan Wiedemann: <milan.wiedemann@phc.ox.ac.uk>

## What are these sessions for?

- Our background research indicated that a lot of you use R...

- But the number of people who say they are competent with it is much lower.

- We want to fix this, the old-fashioned way (no AI here!)

- Unlike the existing R resources here, our sessions are highly targeted.

- We also believe that learning new things is easier when it's collaborative.

## How confident with R are you?

```{r confdiagram}
conf <- read.csv(here::here("confidence.csv"))
hist(conf$confidence, main = NA, col = "lightblue", xlab = "Confidence")
```

## Introduction to GitHub

- We (the team) use GitHub for collaboratively planning and designing these sessions.

- Therefore, that's where you will find all the resources you need.

- You can access our repository here: <https://github.com/milanwiedemann/phc-coding-club>

- Make sure to scroll down far enough to see the README file!

## Demo 1: Plasma effect

```{r plasma}
plasma <- function(x, y) sin(x / 2) + sin(y) + sin((x + y) / 2) + sin(sqrt(x ^ 2 + y ^ 2))
x <- y <- seq(-4 * pi, 4 * pi, length.out = 1000)
bigmat <- outer(x, y, plasma)
image(bigmat, col = hcl.colors(12, "Plasma"), axes = FALSE)
```

## Demo 2: Cuboid

```{r background}
cuboid <- function(w, h, d){
  plot(x = NA,
       xlim = c(0, (w + d) * sqrt(3) / 2),
       ylim = c(0, h + (w + d) / 2),
       xlab = NA,
       ylab = NA,
       asp = 1,
       axes = FALSE)
  
  polygon(x = c(0, w * sqrt(3) / 2, w * sqrt(3) / 2, 0), 
          y = c(w / 2, 0, h, h + w / 2), 
          col = "red")
  
  polygon(x = c(0, w * sqrt(3) / 2, (w + d) * sqrt(3) / 2, d * sqrt(3) / 2), 
          y = c(h + w / 2, h, h + d / 2, h + (w + d) / 2), 
          col = "blue")
  
  polygon(x = c(w * sqrt(3) / 2,
                (w + d) * sqrt(3) / 2,
                (w + d) * sqrt(3) / 2,
                w * sqrt(3) / 2), 
          
          y = c(0, d / 2, h + d / 2, h), 
          col = "yellow")
}

cuboid(2,3,4)

```
